Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-13 numerical.tex
updating

We solve the linear problem by expanding the
perturbations in Chebyshev polynomials $T_l$ up to $l=512$
and discretizing the equations on a grid with
$z\in[-\zmax,\zmax]$. Our standard boundary conditions at the vertical
boundaries is a free surface, 
\begin{align}
  \Delta P \equiv \delta P + \bm{\xi}\cdot\nabla P= 0 \quad \text{at } z=\pm\zmax,
\end{align}
where $\bm{\xi}$ is the Lagrangian displacement, the meridional
components of which are given by $\xi_{x,z} = \ii\delta
v_{x,z}/\sigma$.

@@

Fig. \ref{iso_eigen_kx} compares the numerically obtained eigenvalues 
for the fundamental mode and that calculated analytically from our
analysis of isothermal perturbations in Appendix \ref{iso_discuss}
(Eq. \ref{simple_growth} with $L=1$). We find good agreement in the
wave frequency $\omega$ at all $\khat$ and in growth rates $\nu$ for
$\khat\gtrsim 10$. Growth rates are $O(\epsilon\Omega_k)$, 
consistent with our discussion in Appendix \ref{max_growth1}, but
are over-estimated by our analysis. The match in $\nu$ worsens for
decreasing $\khat$ due to the neglect of global radial structure in
the  analysis. Nevertheless, given the number of approximations made
to calculate growth rates analytically, this agreement is
satisfactory, especially at large $\khat$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_eigen_imag_iso} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  1cm]{figures/compare_eigen_real_iso}
  \caption{Growth rate (top) and real frequency (bottom) of the
    fundamental VSI mode in a disk with $(\gamma,
    \Gamma)=(1.4,1.011)$, $(p,q,\epsilon)=(-1.5,-1,0.1)$ and
    $\beta=10^{-3}$. The solid (dotted) line is obtained numerically
    (analytically).  
    \label{iso_eigen_kx} 
  }
\end{figure}

@@

Fig. \ref{lowfreq_eigenfunc} plots the eigenfunctions $W$ and $\delta 
v_z$ for the fundamental mode with $\khat=10$ (i.e. the mode with
maximum growth rate in Fig. \ref{iso_eigen_kx}). We find $W\propto z$
throughout most of the disk, corresponding to a constant vertical
velocity perturbation, as expected for the fundamental mode according
to explicit solutions developed in Appendix \ref{iso_explicit}. We
show a meridional visualization of this mode in
Fig. \ref{lowfreq_eigenfunc_2d}. Radial velocities are typically much
smaller than vertical velocities. Although there is a constant vertical
velocity throughout the fluid column, most of the momentum is
contained within $\sim 2H$ of the midplane due to the density
stratification.

@@

\begin{figure}
%  \includegraphics[scale=0.345,clip=true,trim=0cm 0cm 2.5cm
%  0cm]{figures/result2d_vel}\includegraphics[scale=0.345,clip=true,trim=1.9cm 0cm 0cm
%  0cm]{figures/result2d_mom} 
  \includegraphics[width=\linewidth]{figures/result2d_iso}
  \caption{Visualization of the fundamental VSI mode in
    Fig. \ref{lowfreq_eigenfunc}. The color scale corresponds to the
    real pressure perturbation $\delta P$ scaled by its maximum value.
    Arrows correspond to the vector field 
    field $\sqrt{\rho}(\delta v_x,\delta v_z)$. Note that the horizontal scale has been scretched for clarity. 
    \label{lowfreq_eigenfunc_2d}
  }
\end{figure}

\subsubsection{Case study: $\khat=10$}
In Fig. \ref{compare_modes_iso_kx10} we plot eigenvalues of unstable
modes found for our standard setup with $\khat=10$. This case is shown
by red crosses. We also plot several other calculations with different
vertical boundary conditions: free surface at $\zmax = 3H$ (black
dots), free surface at $\zmax=7H$ (green triangles) and a rigid
boundary at $\zmax=5H$ (blue cross).

The fundamental mode, marked by `f', corresponds to the eigenvalue
with smallest magnitude $|\sigma|$. We see that the fundamental mode,
while not the most unstable in this case, is more robust against
vertical boundary conditions. Higher order modes with larger growth
rates (i.e. the branch where $\nu$ increases with $|\omega|$) require
larger vertical domains.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_modes_iso_kx10.ps}
  \caption{Unstable modes for a disk with 
    $(\gamma,\Gamma)=(1.4,1.011)$, $(p,q,\epsilon)=(-1.5,-1,0.1)$ and
    $\beta=10^{-3}$. The perturbation wavenumber is $\khat=10$. 
    The standard setup with a free surface ($\Delta P=0$) and 
    vertical domain size $\zmax=5H$ is shown by the red cross. Other
    cases correspond to different boundary condtions applied at
    different heights. The fundamental mode is marked by `f'.   
    \label{compare_modes_iso_kx10} 
  }
\end{figure}

\subsubsection{Comment on `surface modes'} 
Our standard setup with $\khat=10$ in 
Fig. \ref{compare_modes_iso_kx10} (red crosses) differs to that from
\cite{nelson13} and \cite{mcnally14} in the lack of `surface modes' in
our case. These modes lie parallel to the vertical axis in mode
diagrams such as Fig. \ref{compare_modes_iso_kx10}. We find them for 
sufficiently large $\zmax$ and/or $\khat$. % For example, the two
% largest growth rates found for the case with $\zmax=7H$ in
% Fig. \ref{compare_modes_iso_kx10} are surface modes.

To illustrate this, we plot the mode diagram for $\khat=30$ in  
Fig. \ref{compare_modes_iso_kx30} with different boundary 
conditions. Surface modes can be seen with
$\zmax\geq 5H$, examples of which are marked by `s'. These modes have
been discussed in the aforementioned studies. They typically have the
largest growth rates but require sufficiently large vertical domain
size and their eigenvalues vary with $\zmax$.

@@

Fig. \ref{compare_modes_cool_kx10} show unstable modes with $\khat=10$
and $\beta\in[0.01,0.1]$. In addition to our standard setup with 
$\zmax=5H$, we also plot results for $\zmax=3H$ and $\zmax=7H$. In all
three cases the number of unstable modes decrease with increasing
$\beta$. The fundamental mode is not the most unstable for 
$\beta=0.01$ as there is a large number of higher order modes, but
these depend on boundary conditions.

>>>>

We solve the linear problem by expanding the
perturbations in Chebyshev polynomials $T_l$ up to $l=512$
and discretizing the equations on a grid with
$z\in[-\zmax,\zmax]$. Our standard boundary conditions at the vertical
boundaries is a free surface, 
\begin{align}
  \Delta P \equiv \delta P + \bm{\xi}\cdot\nabla P= 0 \quad \text{at } z=\pm\zmax,
\end{align}
where $\bm{\xi}$ is the Lagrangian displacement with meridional 
components $\xi_{x,z} = \ii\delta v_{x,z}/\sigma$.

@@

Fig. \ref{iso_eigen_kx} compares the numerically obtained eigenvalues 
for the fundamental mode and that calculated from 
Eq. \ref{simple_growth} (with $L=1$). We find good agreement in the
wave frequency $\omega$ at all $\khat$ and in growth rates $\nu$ for
$\khat\gtrsim 10$. Growth rates are over-estimated by our analysis but 
are $O(\epsilon\Omega_k)$, which is consistent with the discussion in
Appendix \ref{max_growth1}. The match in $\nu$ worsens for
decreasing $\khat$ due to the neglect of global radial structure in
the analysis. Nevertheless, given the number of approximations made
to calculate growth rates analytically, this agreement is
satisfactory, especially at large $\khat$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/compare_eigen_imag_iso} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  1cm]{figures/compare_eigen_real_iso}
  \caption{Growth rate (top) and real frequency (bottom) of the
    fundamental VSI mode in the fiducial disk model with $\beta =
    10^{-3}$. 
    % a disk with $(\gamma,
    % \Gamma)=(1.4,1.011)$, $(p,q,\epsilon)=(-1.5,-1,0.1)$ and
    % $\beta=10^{-3}$. 
    The solid (dotted) line is obtained numerically
    (analytically).  
    \label{iso_eigen_kx} 
  }
\end{figure}

@@

Fig. \ref{lowfreq_eigenfunc} plots the eigenfunctions $W$ and $\delta 
v_z$ for the fundamental mode with $\khat=10$ (i.e. the mode with
maximum growth rate in Fig. \ref{iso_eigen_kx}). We find $W\propto z$
throughout most of the disk, corresponding to a constant vertical
velocity perturbation, as expected for the fundamental mode according
to explicit solutions developed in Appendix \ref{iso_explicit}. We
show a meridional visualization of this mode in
Fig. \ref{lowfreq_eigenfunc_2d}. The $x$ axis is stretched for
clarity; radial velocities are in fact typically much
smaller than vertical velocities. Most of the meridional kinetic energy is
contained within $\sim 2H$ of the midplane due to the density 
stratification.

@@

\begin{figure}
%  \includegraphics[scale=0.345,clip=true,trim=0cm 0cm 2.5cm
%  0cm]{figures/result2d_vel}\includegraphics[scale=0.345,clip=true,trim=1.9cm 0cm 0cm
%  0cm]{figures/result2d_mom} 
  \includegraphics[width=\linewidth]{figures/result2d_iso}
  \caption{Visualization of the fundamental VSI mode in
    Fig. \ref{lowfreq_eigenfunc}. The color scale corresponds to the
    pressure perturbation $\delta P=\rho W$ scaled by its maximum value.
    Arrows correspond to the vector field field $\sqrt{\rho}(\delta
    v_x,\delta v_z)$. Note that the horizontal axis has been scretched
    for clarity.  
    \label{lowfreq_eigenfunc_2d}
  }
\end{figure}

\subsubsection{Case study: $\khat=10$}
In Fig. \ref{compare_modes_iso_kx10} we plot eigenvalues of unstable
modes with $\khat=10$. The result for our fiducial setup is shown
in black dots. We also plot other calculations with different
vertical boundary conditions: rigid boundary and $\zmax=5H$ (red
cross),  free surface and $\zmax = 3H$ (blue triangles), free surface
and $\zmax=7H$ (green cross).

The fundamental mode, marked by `f', corresponds to the eigenvalue
with the smallest $|\sigma|$. We see that the fundamental mode is
robust against vertical boundary conditions as it appears at the same
position in all cases. Higher order modes with larger growth rates
(i.e. the branch where $\nu$ increases with $|\omega|$) require larger
vertical domains.  The `tail' end with decreasing $\nu$ to the right
of the plot differ significantly between the cases, and are thus
likely associated with boundary conditions.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_modes_iso_kx10.ps}
  \caption{Unstable modes for a disk with 
    $(\gamma,\Gamma)=(1.4,1.011)$, $(p,q,\epsilon)=(-1.5,-1,0.1)$ and
    $\beta=10^{-3}$. The perturbation wavenumber is $\khat=10$. 
    The fiducial setup with a free surface ($\Delta P=0$) and 
    vertical domain size $\zmax=5H$ is shown by black dots Other
    cases correspond to different boundary condtions applied at
    different heights. The fundamental mode is marked by `f'.   
    \label{compare_modes_iso_kx10} 
  }
\end{figure}

\subsubsection{Comment on `surface modes'} 
Our fiducial setup with $\khat=10$ in 
Fig. \ref{compare_modes_iso_kx10} (black dots) differs to that from
\cite{nelson13} and \cite{mcnally14} in the lack of `surface modes'. 
These modes lie almost parallel to the vertical axis in mode 
diagrams such as Fig. \ref{compare_modes_iso_kx10}. We find them for 
sufficiently large $\zmax$ and/or $\khat$. % For example, the two
% largest growth rates found for the case with $\zmax=7H$ in
% Fig. \ref{compare_modes_iso_kx10} are surface modes.

To illustrate this, we plot the mode diagram for $\khat=30$ in  
Fig. \ref{compare_modes_iso_kx30} with different boundary 
conditions. Surface modes are readily identified for 
$\zmax\geq 5H$ and are marked by `s'. These modes have
already been discussed in the aforementioned studies. They typically
have the largest growth rates but require sufficiently large vertical
domain size and their eigenvalues vary with $\zmax$.

@@

Fig. \ref{compare_modes_cool_kx10} show unstable modes with $\khat=10$
and $\beta\in[0.01,0.1]$. In addition to our fiducial setup with 
$\zmax=5H$, we also plot results for $\zmax=3H$ and $\zmax=7H$. In all
three cases the number of unstable modes decrease with increasing
$\beta$. The fundamental mode is not the most unstable for 
$\beta=0.01$ as there is a large number of higher order modes, but
these depend on boundary conditions.
